---
title: ""
date: "`r Sys.Date()`"
output:
  knitrBootstrap::bootstrap_document:
    theme: simplex
    highlight: zenburn
    theme.chooser: TRUE
    highlight.chooser: TRUE
---

```{r knitr_init, echo=FALSE, cache=FALSE, warning=FALSE}
library(knitr)
library(rmdformats)

## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE,
               bootstrap.show.output=FALSE)
opts_knit$set(width=75)
```


R commands can sometimes be rather difficult to follow, so occasionally it can be useful to annotate them with comments. This can be done by typing a hash (#) character: any further text on the same line is ignored by R. Copy the code below onto your Source pane (top-left) and run it while reading the comments.

## Practicals 

These are a few exercises that should help you get acquainted with R. You should calculate or evaluate the R expressions before pressing the "R Output button."

If however, you would like to train by yourself, rather than in the website, [here's a link for the R code]() so you can download it and run in your own computer.

```{r}
# this is a comment: R will ignore it
1 + 1 # this queals to two

#The symbol * means multiply, and ^ means "to the power", so this gives 5 times (10 squared), i.e. 500
5*10^2

# A different function: "sqrt" takes a single argument, returning its square root.
sqrt(25)

# The result of a function can be used as part of a further analysis
sqrt(25-9)+100        

# For example, let's learn about the function max()
# This function returns the maximum value of all its arguments
max(-10, 0.2, 4.5)  

# Now you can use results of functions as arguments to other functions
sqrt(2 * max(-10, 0.2, 4.5))             

# The log() function returns the logarithm of its first argument
log(100)

# By default this is the natural logarithm (base "e")
log(2.718282)         

# But you can change the base of the logarithm using the "base" argument
log(100, base=10)
log10(100)

# R knows about infinity (and minus infinity)
1/0
-1/0

# undefined results take the value NaN ("not a number")
0/0

# for the mathematically inclined, you can force R to use complex numbers
sqrt(0i-9)

# R has another special symbol for 'empty'
sum(c(1, 2, NULL, 4))
length(c(1, 2, NULL, 4))
```



